---
title: "6: Part 1 - Generalized Linear Models"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2021"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Describe the components of the generalized linear model (GLM)
2. Apply special cases of the GLM (linear regression) to real datasets
3. Interpret and report the results of linear regressions in publication-style formats
3. Apply model selection methods to choose model formulations

## Generalized Linear Models (GLMs)

The analysis of variance (ANOVA), and linear regression are special cases of the **generalized linear model** (GLM). The GLM also includes analyses not covered in this class, including logistic regression, multinomial regression, chi square, and log-linear models. The common characteristic of general linear models is the expression of a continuous response variable as a linear combination of the effects of categorical or continuous explanatory variables, plus an error term that expresses the random error associated with the coefficients of all explanatory variables. The explanatory variables comprise the deterministic component of the model, and the error term comprises the stochastic component of the model. Historically, artificial distinctions were made between linear models that contained categorical and continuous explanatory variables, but this distinction is no longer made. The inclusion of these models within the umbrella of the GLM allows models to fit the main effects of both categorical and continuous explanatory variables as well as their interactions. 

Framework:
Data(continuous variable y) = model(one or more variables (x[n] with x being continuous or categorical)) + residual error term
*GLM is based on the assumption that the data residuals approximate a normal distribution

### Choosing a model from your data: A "cheat sheet"

**One-way ANOVA (Analysis of Variance):** Continuous response, one categorical explanatory variable with more than two categories.

**Two-way ANOVA (Analysis of Variance)** Continuous response, two categorical explanatory variables.

**Single Linear Regression** Continuous response, one continuous explanatory variable.

**Multiple Linear Regression** Continuous response, two or more continuous explanatory variables.

**ANCOVA (Analysis of Covariance)** Continuous response, categorical explanatory variable(s) and  continuous explanatory variable(s).

If multiple explanatory variables are chosen, they may be analyzed with respect to their **main effects** on the model (i.e., their separate impacts on the variance explained) or with respect to their **interaction effects,** the effect of interacting explanatory variables on the model. 

### Assumptions of the GLM

The GLM is based on the assumption that the data residuals approximate a normal distribution (or a linearly transformed normal distribution). We will discuss the non-parametric analogues to several of these tests if the assumptions of normality are violated. For tests that analyze categorical explanatory variables, the assumption is that the variance in the response variable is equal among groups. Note: environmental data often violate the assumptions of normality and equal variance, and we will often proceed with a GLM even if these assumptions are violated. In this situation, justifying the decision to proceed with a linear model must be made.

#### Hypothesis testing:

Conduct hypothesis testing to analayze evidence provided by the data and to make decisions.
A statistical hypothesis is an assumption about a population parameter that may or may not be true (i.e., mean, variance, etc).
Hypothesis testing usually has: Ho (null hypothesis) and H1 (alternative hypothesis)

1. State the hypothesisand identify the claim
2. Find the critical values fromt the appropriate table
3. Compute the test value
4. Make the decision to reject or not the null hypothesis (if p-value less than or equal to alpha, reject; if p-value greater than alpha, do not reject).Alpha is confidence level.

#### Simple Linear Regression

Regression - technique for fitting a line to a set of data points.

Simple liner regression - simplest; involves a linear relationship between 2 variables. The object of a simple linear regression is to obtain an equation of a straight line that minimizes the sum of squared vertical deviations from the line (i.e., the least squares criterion).

#### Standard Error

Standard error of estimate is a measure of the scatter of points around a regression line.

If the standard error is relatively small, the predictions using the linear equation will tend to be more accurate than if the standard error is larger.

#### Correlation Coefficient

The correlation coefficient (r) is the measure of the strength and the direction of the relationship between two variables. Ranges between -1.0 and +1.0. The r^2 is a measure of the percentage of variability in the values of y that is "explained" by the independent variable. This ranges between 0 and 1.0.

#### Residuals

After fitting a regression model, check the residual plots first to be sure that you have unbiased estimates.

##### P-values and coefficients in regression analysis

The p-values for the coefficients indicate whether these relationships are statistically significant (determines whether the relationships you observe in sample also exist in larger population). The p-value for each independent variable tests the null hypothesis (Ho: independent variable has no correlation with the dependent variable).

If the p-value for a variable is less than your significance level, your sample data provide enough evidence to reject the null hypothesis for the entire population. 


## Set up
```{r, message = FALSE}
getwd()
library(tidyverse)
options(scipen = 4)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Linear Regression
A linear regression is comprised of a continuous response variable, plus a combination of 1+ continuous response variables (plus the error term). The deterministic portion of the equation describes the response variable as lying on a straight line, with an intercept (alpha) and a slope (beta) term. Also includes error term (epsilon) which is the distance of y value to true value of y on line. The equation is thus a typical algebraic expression: 
$$ y = \alpha + \beta*x + \epsilon $$

The goal for the linear regression is to find a **line of best fit**, which is the line drawn through the bivariate space that minimizes the total distance of points from the line. This is also called a "least squares" regression. The remainder of the variance not explained by the model is called the **residual error.** 

The linear regression will test the null hypotheses that

1. The intercept (alpha) is equal to zero.
2. The slope (beta) is equal to zero

Whether or not we care about the result of each of these tested hypotheses will depend on our research question. Sometimes, the test for the intercept will be of interest, and sometimes it will not.

Important components of the linear regression are the correlation and the R-squared value. The **correlation** is a number between -1 and 1, describing the relationship between the variables. Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables. 

## Simple Linear Regression

*Regressing one variable y on one single variable x

For the NTL-LTER dataset, can we predict irradiance (light level) from depth?

```{r}
irradiance.regression <- lm(PeterPaul.chem.nutrients$irradianceWater ~ PeterPaul.chem.nutrients$depth)
#lm is linear model; need to provide y (irradiance) and x(depth)

# another way to format the lm() function - that way don't need $ to call columns
irradiance.regression <- lm(data = PeterPaul.chem.nutrients, irradianceWater ~ depth)
#tilda is how model is written; irradiance is y (dependent variable) an depth is x
summary(irradiance.regression)
#results: slope (-95.8) shows there is a negative relationship. P-value less than .05, showing that they are statistically different. So worthwhile to determine relationship. Degrees of freedom relate to number of samples. R^2 explains that depth explains about 30% difference in irradiance.

# Correlation
cor.test(PeterPaul.chem.nutrients$irradianceWater, PeterPaul.chem.nutrients$depth) #strong correlation at 0.55
```
Question: How would you report the results of this test (overall findings and report of statistical output)?

>  

So, we see there is a significant negative correlation between irradiance and depth (lower light levels at greater depths), and that this model explains about 31 % of the total variance in irradiance. Let's visualize this relationship and the model itself. 

An exploratory option to visualize the model fit is to use the function `plot`. This function will return four graphs, which are intended only for checking the fit of the model and not for communicating results. The plots that are returned are: 

1. **Residuals vs. Fitted.** The value predicted by the line of best fit is the fitted value, and the residual is the distance of that actual value from the predicted value. By definition, there will be a balance of positive and negative residuals. Watch for drastic asymmetry from side to side or a marked departure from zero for the red line - these are signs of a poor model fit.

2. **Normal Q-Q.** The points should fall close to the 1:1 line. We often see departures from 1:1 at the high and low ends of the dataset, which could be outliers. 

3. **Scale-Location.** Similar to the residuals vs. fitted graph, this will graph the squared standardized residuals by the fitted values. 

4. **Residuals vs. Leverage.** This graph will display potential outliers. The values that fall outside the dashed red lines (Cook's distance) are outliers for the model. Watch for drastic departures of the solid red line from horizontal - this is a sign of a poor model fit.

```{r, fig.height = 3, fig.width = 4}
par(mfrow = c(2,2), mar=c(4,4,4,4)) #par is dividing cells into a grid (4 cells, 2x2 grid; mar finds limits of grid (size of plots)
plot(irradiance.regression) #provide plot with linear model object
#first plot is fitted values vs residuals - want symmetry with red line and be as flat as possible
#nomral QQ plot - dashed line responds to QQ for normal distribution; circles relate to residuals. Want circles to relate to dashed line as much as possible. Larger and smaller n's can cause deviation away from dashed line (may due to outliers, not that it doesn't follow normal distribution)
#fitted values vs sq. residuals - looking for symmetry to red line
#residuals vs legerage - outliers outside of dashed lines is outliers of model; drastic departures are sign of poor model fit
par(mfrow = c(1,1))
```

The option best suited for communicating findings is to plot the explanatory and response variables as a scatterplot. 

```{r, fig.height = 3, fig.width = 4}
# Plot the regression
irradiancebydepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  ylim(0, 2000) + #changing plotting window to show values between 0 and 2000
  geom_point() 
print(irradiancebydepth) 
#no linear relationship between x and y - so need to log transform
```

Given the distribution of irradiance values, we don't have a linear relationship between x and y in this case. Let's try log-transforming the irradiance values. Note we also removing the observations that seems to ba an outlier.

```{r, fig.height = 3, fig.width = 4}
#want to log of data to see if there is a linear relationships. Need to eliminate zeros and outlier
PeterPaul.chem.nutrients <- filter(PeterPaul.chem.nutrients, 
                                   irradianceWater != 0 & irradianceWater < 5000)
irradiance.regression2 <- lm(data = PeterPaul.chem.nutrients, log(irradianceWater) ~ depth)
summary(irradiance.regression2)
#R^2 increased so indicates this is a better model of the data

par(mfrow = c(2,2), mar=c(4,4,4,4))
plot(irradiance.regression2)
par(mfrow = c(1,1))

# Add a line and standard error for the linear regression
irradiancebydepth2 <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  geom_smooth(method = "lm") +
  scale_y_log10() + #take log - y axis is log transformed, includes confidence interval
  geom_point() 
print(irradiancebydepth2) 

# SE - confidence interval around smooth can also be removed
irradiancebydepth2 <- 
    ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
    geom_point() +
    scale_y_log10() +  #log transformed values for irradianceWater
    geom_smooth(method = 'lm', se = FALSE, color = "black")
print(irradiancebydepth2)

# Make the graph attractive

```

## Multiple Linear Regression
It is possible, and often useful, to consider multiple continuous explanatory variables at a time in a linear regression. For example, total phosphorus concentration in Paul Lake (the unfertilized lake) could be dependent on depth and dissolved oxygen concentration: 

``` {r, fig.height = 3, fig.width = 4}
TPregression <- lm(data = subset(PeterPaul.chem.nutrients, lakename == "Paul Lake"), 
                   tp_ug ~ depth + dissolvedOxygen) #looking at the two variables and dissolved oxygen is residual variable
summary(TPregression)

TPplot <- ggplot(subset(PeterPaul.chem.nutrients, lakename == "Paul Lake"), 
                 aes(x = dissolvedOxygen, y = tp_ug, color = depth)) +
  geom_point() +
  xlim(0, 20)
print(TPplot)

par(mfrow = c(2,2), mar=c(4,4,4,4))
plot(TPregression)
par(mfrow = c(1,1))
#want red line to be flat for residuals vs fitted - there is some pattern so not a good fit
#QQ plot ok
#outliers in residuals vs leverage
```

## Correlation Plots
We can also make exploratory plots of several continuous data points to determine possible relationships, as well as covariance among explanatory variables. 

```{r, fig.height = 3, fig.width = 4}
#install.packages("corrplot")
#create correlation matrix -  see all correlations among variables in data set
library(corrplot)
PeterPaulnutrients <- 
  PeterPaul.chem.nutrients %>%
  select(tn_ug:po4) %>%
  na.omit()
PeterPaulCorr <- cor(PeterPaulnutrients) #create correlation matrix
corrplot(PeterPaulCorr, method = "ellipse") #create correlation plot; oval shapes represent one scatter plot of one variable against the other; the thinner the scatter plot, the higher the correlation (can approximate by straight line); red is strong but neg correlation, blue is strong correlation; color and shape of ellipses give strength and sign of correlation; symmetric matrix
corrplot.mixed(PeterPaulCorr, upper = "ellipse") #mixed is mix of numbers and ellipse; shows exactly the value is with the ellipses
```

## AIC to select variables

However, it is possible to over-parameterize a linear model (more variables that you need). Adding additional explanatory variables takes away degrees of freedom, and if explanatory variables co-vary the interpretation can become overly complicated. Remember, an ideal statistical model balances simplicity and explanatory power! To help with this tradeoff, we can use the **Akaike's Information Criterion (AIC)** to compute a stepwise regression that either adds explanatory variables from the bottom up or removes explanatory variables from a full set of suggested options. The smaller the AIC value, the better. 

Let's say we want to know which explanatory variables will allow us to best predict total phosphorus concentrations. Potential explanatory variables from the dataset could include depth, dissolved oxygen, temperature, PAR, total N concentration, and phosphate concentration.

```{r}
Paul.naomit <- PeterPaul.chem.nutrients %>%
  filter(lakename == "Paul Lake") %>%
  na.omit()

TPAIC <- lm(data = Paul.naomit, tp_ug ~ depth + dissolvedOxygen + 
              temperature_C + tn_ug + po4)

#Choose a model by AIC in a Stepwise Algorithm
step(TPAIC)
TPmodel <- lm(data = Paul.naomit, tp_ug ~ dissolvedOxygen + temperature_C + tn_ug)
summary(TPmodel)

#want to have the lowest AIC to have the regression that best fits the data (considering oxygen, nitrogen, and temp)
#step function also gives coefficients of 3 variables


```