---
title: "7: Time Series"
author: "Environmental Data Analytics | Adapted by John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2021"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Discuss the purpose and application of time series analysis for environmental data
2. Explore the components of times series: trend, seasonal, random
3. Learn how to perform stationarity test

## Set up

Today we will work with two datasets. The USGS dataset on discharge at the Eno River and a new dataset we haven't explored yet on wind speed. The data file is available at "./Data/Raw/Wind_Speed_PortArthurTX.csv". It contains average wind speed in monthly time steps (elevation = 5 meters). The data is available from NOAA National Centers for Environmental Information (NCEI) [here][https://www.ncdc.noaa.gov/cdo-web/datasets#GSOM].

```{r, message = FALSE}

library(tidyverse)
library(lubridate)
#install.packages("trend") - runs stationary tests
library(trend)
#install.packages("zoo") - replaces missing values in time series
library(zoo)
#install.packages("Kendall") - specific to one stationary test to be performed
library(Kendall)
#install.packages("tseries")
library(tseries)

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

#Read Eno river data
EnoDischarge <- read.csv("./Data/Processed/USGS_Site02085000_Flow_Processed.csv",
                         stringsAsFactors = TRUE)
EnoDischarge$datetime <- as.Date(EnoDischarge$datetime, format = "%Y-%m-%d")

#Read wind speed data
wind_data <- read.csv(file="./Data/Raw/Wind_Speed_PortArthurTX.csv",header=TRUE,
                      stringsAsFactors = TRUE)
wind_data$DATE <- ym(wind_data$DATE)

```


## Time Series Analysis overview

Time series are a special class of dataset, where a response variable is tracked over time. The frequency of measurement and the timespan of the dataset can vary widely. At its most simple, a time series model includes an explanatory time component and a response variable. Mixed models can include additional explanatory variables (check out the `nlme` and `lme4` R packages). We will cover a few simple applications of time series analysis in these lessons, with references for how to take analyses further.

Can be a discrete or continuous time series - is a function of time - defined by values of y1,y2 of a variable y at times t1,t2:

y=f(t)

Time frame dependent on data set that is available (short, medium, long-term)

### Trend
- general tendency to grow or decline over a long period
-easiest to detect
-maybe linear or non linear

### Cycle
-up and down repetitive movement
-repeat itself over a long period of time
example: business cycle(prosperity, decline,depression, recovery)

### Seasonal Variation
-up and down repetitive movement occuring periodically (short duration)
-can be caused by climate, weather condition, or custom traditions or habits

### Random variations
-erratic movements that are not predictable because they do not follow any pattern
-example: strike, fire, flood, earthquake, etc

### Terms:
-stationary data - time series variable with no significant downward or upward trend over time
-nonstationary data - time eseries that has significant upward/downward trend
-seasonal data - time series exhibiting repeating patterns at regular intervals over time

### Autocorrelation 

Measure of dependence between 2 adjacent values of the same variables (auto=self correlation)

How to compute:
-in the context of a single variable, yt is the original series and ys is a lagged version of the series (same variable as yt, but shift above original/previous observation)
Pt,s = Corr(yt,ys)
*think of placing in excel and shifting one cell to create a lagged version

Main conclusion: autocovariance and autocorrelation function five information about the dependence structure of a time series
-if observation found at time is significant, it is dependent on the observation found at t-1

### Stationary Process

The basic idea of stationary is that the probability laws that govern the behavior of the process do not change over time
-the mean, standard deviation, autocorrelation do not change over time (always have same parameters no matter what time)
The distribution of observations at these points = The distribution of observations at these points
-example: white noise series - series of independent, identically distributed (i.i.d) random variables {et};{et} is a stationary process. Assume that white noise series has a mean=0 and Var (et) = sigma^2[e]

### Partial Autocorrelation Function

The ACF of a stationary process yt at lag h: 

pt,t-h = Corr(yt,yt-h)

measures the linear dependency among the process variables yt and yt-h, but the dependency structure among the intermediate variables:
yt,yt-1,yt-2,yt-3...yt-h+2,yt-h+1,yt-h
*intermediate variables are those between yt and yt-h
also plays an important role on the value ACF

*removes influence of all intermediate variables - when removed, you would only have the direct correlation between yt and yt-h (PACF = partial ACF)

-The ACF and PACF measure the temporal dependency of a stochastic process
-You will always build the ACFand PACF before fitting a model to a stochastic process
-The ACF and PACF give us information about the regressive component of the series (regression on previous observations,no external variables, only lagged variables)

### Trend Component

Long-term tendency: increase (upward movmenet) or decrease (downward movement)
Trend can be linear or nonlinear

Linear trend can be written as yi = beta0 + beta1ti + epsiloni

Slope (beta1) and the intercept (beta0) are the unknown parameters, and epsiloni is the error term (the error or resudual is the distance from point yi to the residual)

Want to remove trend because want to model a stationary data set (same mean over time that is easier to model than one that has a trend).Need to create a new variable and for each observation of y, need to subtract estimated trend
ydetrendt = yt - (beta0 -beta1t)

### Seasonal Component

Short term regular wave-like patterns. Observed within 1 year, often monthly or quarterly.
To estimate a seasonal trend, asuume the observed series can be represented as:

yt=mut + xt

where e[xt] = 0

### Deterministic Trend

Detrending is accomplished by running a regression and obtaining a series of residuals. The residuals will give the detrended series (trend-stationarity)

Stochastic trend has different behaviors over time (difference stationarity). Although trend-stationarity and difference-stationarity are both trending over time, the stationarity is achieved by a distinctive procedure. In the case of difference-stationarity, stationarity is achieved by differencing the series. Sometimes we need to difference the series more than once.

### Stationarity tests

-Mann-Kendall test cannot be applied to seasonal data - commonly apploed to deterministic trends in series of environmental data, climate data, or hydrological data. Hypothesis test - null is that it is stationarity. Main test statistic is S - comparison of observations from sample. Want to see if future values are bigger. Will check magnitude of S and its significance based on observations (bigger number of observations,bigger S will be in order to significant).
Use package "Kendall" and function MannKendall() and for seasonal data, use SeasonalMannKendall()

-Spearman's rank correlation coefficient - a statistical measure of the strength of a monotonic relationship. Unlike Pearson's the relationship, it can check if a relationship is not linear (deos not matter the proportional increase). Because it is more general, it is always going to be higher than Pearson's. Need to verify there is a monotonic trend by computing the spearman correlation between data and series T. If correlation is close to zero than there is no trend. Use function cor() cor.test() from package "stats". Cor.test will include the coefficient which will tell you the significance of the test (related to sample size and strength of correlation).

-Dick-Fuller test - tests a unit root (checks for a stochastic trend). Null hypothesis is that it contains a unit root (hard to control, needs treatments before applying time series model), whereas the alternative is stationary. In order to remove a unit root need to differentiate series. More general case can include more lags, need to use Augmented Dickey-Fuller (ADF) test. From package "tseries" and use function adf.test


### Opportunities

Analysis of time series presents several opportunities. For environmental data, some of the most common questions we can answer with time series modeling are:

* Has there been an increasing or decreasing **trend** in the response variable over time?
* Can we **forecast** conditions in the future?

### Challenges

Time series datasets come with several caveats, which need to be addressed in order to effectively model the system. A few common challenges that arise (and can occur together within a single dataset) are: 

* covariance and correlation measure joint variability of 2 variables; measure of linear dependence of 2 variables

* **Autocorrelation**: Data points are not independent from one another (i.e., the measurement at a given time point is dependent on previous time point(s))
* **Data gaps**: Data are not collected at regular intervals, necessitating *interpolation* between measurements.
* **Seasonality**: seasonal patterns in variables occur at regular intervals, impeding clear interpretation of a monotonic (unidirectional) trend.
* **Heteroscedasticity**: The variance of the time series is not constant over time
* **Covariance**: the covariance of the time series is not constant over time

## Handling data gaps and missing data. Example: Eno River Discharge

River discharge is measured daily at the Eno River gage station. Since we are working with one location measured over time, this will make a great example dataset for time series analysis. 

Let's look at what the dataset contains for mean daily discharge.

```{r}
ggplot(EnoDischarge, aes(x = datetime, y = discharge.mean)) +
  geom_line() +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)"))
```

Notice there are missing data from 1971 to 1985. Gaps this large are generally an issue for time series analysis, as we don't have a continuous record of data or a good way to characterize any variability that happened over those years. We will illustrate a few workarounds to address these issues. 

Let's start by removing the NAs and splitting the dataset into the early and late years. 

```{r}
EnoDischarge.complete <- EnoDischarge %>%
  drop_na(discharge.mean)

EnoDischarge.early <- EnoDischarge.complete %>%
  filter(datetime < as.Date("1985-01-01"))

EnoDischarge.late <- EnoDischarge.complete %>%
  filter(datetime > as.Date("1985-01-01"))
```

## Decomposing a time series dataset

A given time series can be made up of several component series: 

1. A **seasonal** component, which repeats over a fixed known period (e.g., seasons of the year, months, days of the week, hour of the day)
2. A **trend** component, which quantifies the upward or downward progression over time. The trend component of a time series does not have to be monotonic.
3. An **error** or **random** component, which makes up the remainder of the time series after other components have been accounted for. This component reflects the noise in the dataset. 
4. (optional) A **cyclical** component, which repeats over periods greater than the seasonal component. A good example of this is El Niño Southern Oscillation (ENSO) cycles, which occur over a period of 2-8 years.

### Example: Eno discharge

We will decompose the EnoDischarge.late data frame for illustrative purposes today. It is possible to run time series analysis on detrended data by subtracting the trend component from the data. However, detrending must be done carefully, as many environmental data are bounded by zero but are not treated as such in a decomposition. If you plan to use decomposition to detrend your data, please consult time series analysis guides before proceeding.

We first need to turn the discharge data into a time series object in R. This is done using the `ts` function. Notice we can only specify one column of data and need to specify the period at which the data are sampled. The resulting time series object cannot be viewed like a regular data frame.

Note: time series objects must be equispaced. In our case, we have daily data with no NAs in the data frame, so we don't need to worry about this. We will cover how to address data that are not equispaced later in the lesson.

```{r}
EnoDischarge.late_ts <- ts(EnoDischarge.late$discharge.mean, start = c(1985,1), frequency = 365)
#frequency 365 - daily data; reduces day by one
```

The `stl` function decomposes the time series object into its component parts. We must specify that the window for seasonal extraction is either "periodic" or a specific number of at least 7. The decomposition proceeds through a loess (locally estimated scatterplot smoothing) function.

```{r}
?stl #seasonal decomposition time series by Loess (break into 3 components); first argument will always be univariate time series to be decomposed (need to run each decomposition on each variable separately); s.windo is 2nd argument and is seasonal window (can be periodic - whatever frequency you specified when creating time series object) or just use the pan (in lags)

# Generate the decomposition
EnoDischarge.late_Decomposed <- stl(EnoDischarge.late_ts, s.window = "periodic")

# Visualize the decomposed series. 
plot(EnoDischarge.late_Decomposed)

# We can extract the components and turn them into data frames
EnoDischarge.late_Components <- as.data.frame(EnoDischarge.late_Decomposed$time.series[,1:3])
#can use $ after time series data frame to view series; the three colums add to be the observed data

EnoDischarge.late_Components <- mutate(EnoDischarge.late_Components,
        Observed = EnoDischarge.late$discharge.mean,     
        Date = EnoDischarge.late$datetime)
#add 2 columns, grabbing info, and adding day time information

# Visualize how the trend maps onto the data
ggplot(EnoDischarge.late_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Discharge (ft"^3*"/s)"))

# Visualize how the seasonal cycle maps onto the data
ggplot(EnoDischarge.late_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Discharge (ft"^3*"/s)"))

```

Note that the decomposition can yield negative values when we apply a seasonal adjustment or a trend adjustment to the data. The decomposition is not constrained by a lower bound of zero as discharge is in real life. Make sure to interpret with caution!


## Trend analysis

Two types of trends may be present in our time series dataset: **monotonic/deterministic** or **stochastic**. Monotonic trends are a gradual shift over time that is consistent in direction, for example in response to land use change. 

A third type of trend we haven't talked about is the **step** trend, also knows as a level shift. Step trends are a distinct shift at a given time point, for example in response to a policy being enacted. 


### Monotonic trend analysis

In general, detecting a monotonic trend requires a long sequence of data with few gaps. If we are working with monthly data, a time series of at least five years is recommended. Gaps can be accounted for, but a gap that makes up more than 1/3 of the sampling period is generally considered the threshold for considering a gap to be too long (a step trend analysis might be better in this situation). 

Adjusting the data may be necessary to fulfill the assumptions of a trend test. A common method to replace missing values is **interpolation**. Common interpolation methods: 

* **Piecewise constant**: also known as a "nearest neighbor" approach. Any missing data are assumed to be equal to the measurement made nearest to that date (could be earlier or later).
* **Linear**: could be thought of as a "connect the dots" approach. Any missing data are assumed to fall between the previous and next measurement, with a straight line drawn between the known points determining the values of the interpolated data on any given date.
* **Spline**: similar to a linear interpolation except that a quadratic function is used to interpolate rather than drawing a straight line.

#### Example: interpolation

The Eno River discharge data doesn't have any short periods of missing data, so interpolation would not be a good choice for that dataset. We will illustrate a linear interpolation using the wind speed dataset. 


```{r}
head(wind_data)
summary(wind_data$AWND)

# Adding new column with no missing obs, just for illustration purpose
# In real applications you will simply replace NAs
wind_data_clean <- 
  wind_data %>% 
  mutate( AWND.clean = zoo::na.approx(AWND) )
#package::specific function - don't need to download package as calls function

summary(wind_data_clean$AWND.clean)
#Note the NA is gone
#interpolation methods won't change main properties of distribution (mean,min,max)

ggplot(wind_data_clean ) +
  geom_line(aes(x = DATE, y = AWND.clean), color = "red") +
  geom_line(aes(x = DATE, y = AWND), color = "black") +
  ylab("Average Wind Speed") 
#need to replace NA with interpolation when in middle - need na.approx, Can only use na.omit if NAs at beginning or end of time series
```

### Monotonic trend analysis, continued

Specific tests for monotonic trend analysis are listed below, with assumptions and tips: 

* **linear regression**: no seasonality, fits the assumptions of a parametric test. Function: `lm`
* **Mann-Kendall**: no seasonality, non-parametric, missing data allowed. Function: `MannKendall()` (package: Kendall)
* **Seasonal Mann-Kendall**: seasonality, non-parametric `SeasonalMannKendall` (package: Kendall)
* **Spearman Rho**: no seasonality, non-parametric, missing data allowed. Function: `cor.test(method="spearman")` (package: stats)

Specific test for stochastic trend analysis:

* **Augmented Dickey Fuller**: no seasonality, non-parametric, missing data not allowed. Function: `adf.test()` (package: tseries)


#### Example: monotonic trend analysis

Let's refer to our wind speed data. We already performed interpolation, but we still need to create our time series object and decompose the series to find out which stationarity test we can apply. 

Note that wind speed has a seasonal cycle. We might be interested in knowing how (if) speed has changed over the course of measurement while incorporating the seasonal component. In this case, we will use a Seasonal Mann-Kendall test to figure out whether a monotonic trend exists.


```{r}

# Generate time series (trend test needs ts, not data.frame)
f_month <- month(first(wind_data_clean$DATE))
f_year <- year(first(wind_data_clean$DATE))
wind_data_ts <- ts(wind_data_clean$AWND.clean,
                   start=c(f_year,f_month),
                   frequency=12) 

#decompose
wind_data_decomp <- stl(wind_data_ts,s.window = "periodic")
plot(wind_data_decomp)

# Run SMK test
wind_data_trend1 <- Kendall::SeasonalMannKendall(wind_data_ts)

# Inspect results
wind_data_trend1
summary(wind_data_trend1)
#null hypothesis is that data is stationary. however if p-value less than 0.05, reject null because have a trend


wind_data_trend2 <- trend::smk.test(wind_data_ts)
# Inspect results
wind_data_trend2
summary(wind_data_trend2)
#SMK shows results of each value

#Visualization
wind_data_plot <-
ggplot(wind_data, aes(x = DATE, y = AWND)) +
  geom_point() +
  geom_line() +
  ylab("Average Wind Speed") +
  geom_smooth( method = lm )
print(wind_data_plot)

```

What would we conclude based on these findings? 

> Answer:


## Forecasting with Autoregressive and Moving Average Models (ARMA)

We might be interested in characterizing a time series in order to understand what happened in the past and to effectively forecast into the future. Two common models that can approximate time series are **autoregressive** and **moving average** models. To classify these models, we use the  **ACF (autocorrelation function)** and the **PACF (partial autocorrelation function)**, which correspond to the autocorrelation of a series and the correlation of the residuals, respectively. 

**Autoregressive** models operate under the framework that a given measurements is correlated with  previous measurements. For example, an AR1 formulation dictates that a measurement is dependent on the previous measurement, and the value can be predicted by quantifying the lag. 

**Moving average** models operate under the framework that the covariance between a measurement and the previous measurement is zero. While AR models use past forecast *values* to predict future values, MA models use past forecast *errors* to predict future values.

Here are some great resources for examining ACF and PACF lags under different formulations of AR and MA models. 
https://nwfsc-timeseries.github.io/atsa-labs/sec-tslab-autoregressive-ar-models.html
https://nwfsc-timeseries.github.io/atsa-labs/sec-tslab-moving-average-ma-models.html

ARMA models require stationary data. This means that there is no monotonic trend over time and there is also equal variance and covariance across the time series. The function `adf.test` will determine whether our data are stationary. The null hypothesis is that the data are not stationary, so we infer that the data are stationary if the p-value is < 0.05.

While some processes might be easy to identify, it is often complicated to predict the order of AR and MA processes. To get around this issue, it is often necessary to run multiple potential formulations of the model and see which one results in the most parsimonious fit using AIC. The function `auto.arima` does this automatically.

